\documentclass{article}

\usepackage{Style}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{tikz}
\usepackage{listings}
\usepackage{color}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{tcolorbox}
\usepackage{qrcode}

% Python code highlighting
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstset{style=mystyle}

\title{Structural Optimization Techniques}
\author{Bilal TAYFUR}

\begin{document}

\begin{titlepage}
\thispagestyle{empty}
\maketitle

\begin{abstract}
This course provides a comprehensive introduction to the optimization of structural systems. Starting from the fundamentals of optimization theory, students will learn the general difference between classical and modern optimization methods and gain the ability to apply these methods in the design and analysis of structural systems. 

The codes of the relevant example can be accessed in the GitHub repository of the course via the QR codes given on the right side of the topics. This repo is updated every year. Therefore, current lecture notes and newly added applications can also be accessed through the repo and can be contributed if desired.

\vspace{10pt}
\begin{center}
\qrcode[height=4cm]{https://github.com/btayfur/structural-optimization}
\end{center}



\vspace{10pt}
\noindent The following books are recommended for students who wish to explore the topics in more depth:

\begin{itemize}
    \item Cottle, R. W., \& Thapa, M. N. (2017). \textbf{Linear and Nonlinear Optimization}. Springer.
    \item Rao, S. S. (2019). \textbf{Engineering Optimization: Theory and Practice, 5th Edition}. John Wiley \& Sons.
    \item Arora, J. S. (2016). \textbf{Introduction to Optimum Design, 4th Edition}. Academic Press.
    \item Yang, X. S. (2021). \textbf{Nature-Inspired Optimization Algorithms, 2nd Edition}. Elsevier.
    \item Bendsøe, M. P., \& Sigmund, O. (2003). \textbf{Topology Optimization: Theory, Methods, and Applications}. Springer.
\end{itemize}

\end{abstract}

\setcounter{tocdepth}{3}
\tableofcontents
\end{titlepage}

\newgeometry{top=20mm,bottom=25mm,right=80mm,left=20mm}

% Including weekly contents
\input{weeks_new/week01}
\input{weeks_new/week02}
\input{weeks_new/week03}
\input{weeks_new/week04}
\input{weeks_new/week05}
\input{weeks_new/week06}
\input{weeks_new/week07}
\input{weeks_new/week08}
\input{weeks_new/week09}
\input{weeks_new/week10}
\input{weeks_new/week11}
\input{weeks_new/week12}
\input{weeks_new/week13}
%\input{weeks_new/week14}

\end{document} 